---
title: "4 – Solutions"
subtitle: "Statistics Data Analysis with R"
author: "Clemens Brunner"
date: 2025-03-24
format:
  html:
    page-layout: full
engine: knitr
knitr:
  opts_chunk: 
    R.options:
      width: 120
highlight-style: github
title-block-banner: true
theme:
  light: flatly
  dark: darkly
---

## Exercise 1

```{r}
u = seq(98, 50, -2)
v = seq(0, 48, 2)
length(u)  # 25 elements -> 5 rows require 5 columns
length(v)  # 25 elements -> 5 rows require 5 columns
dim(u) = c(5, 5)
dim(v) = c(5, 5)
(r = cbind(u, v))
class(r)  # matrix
```


## Exercise 2

```{r}
dim(r)  # 5 rows, 10 columns
length(r)  # 50 elements
r[4, 6]  # 6
rowMeans(r)
colMeans(r)
mean(r[3:5, 1:2])  # 87
```


## Exercise 3

```{r}
df = data.frame(
    name=c(
        "Ben", "Emma", "Luis", "Mia", "Paul",
        "Hanna", "Lukas", "Sophia", "Jonas", "Emilia"
    ),
    gender=rep(c("m", "f"), 5),
    value=c(11, 76, 42, 8, 32, 96, 88, 65, 14, 50)
)
df
str(df)  # character, character, numeric
```


## Exercise 4

```{r}
df_f = df[df[, "gender"] == "f", ]
df_f
```

:::{.callout-note}
There are additional ways to create `df_f`:

```{r}
df_f = df[df$gender == "f", ]
df_f = df[df[["gender"]] == "f",]
df_f = df[df[[2]] == "f",]
df_f = df[df[, 2] == "f",]
```
:::

Here are some ways to access a the first column:

```{r}
df_f[, 1]  # column 1
df_f[, "name"]  # column "name"
df_f$name  # column "name"
df_f[["name"]]  # column "name"
df_f[[1]]  # column 1
```


## Exercise 5

```{r}
dim(mtcars)  # 32 rows, 11 columns
mtcars1 = mtcars[mtcars$mpg > 25, ]
dim(mtcars1)  # 6 rows, 11 columns
```


## Exercise 6

```{r}
library(tibble)

air = as_tibble(airquality)
```

```{r}
airquality[, 1]
air[, 1]
airquality[, "Ozone"]
air[, "Ozone"]
```

Extracting a column with square brackets returns a vector for data frames, but a tibble for tibbles.

```{r}
airquality$Ozone
air$Ozone
airquality[["Ozone"]]
air[["Ozone"]]
```

These variants always return the data in the column as a vector.

Tibbles are more consistent when indexing with square brackets because the result is always a tibble. For data frames, the result is sometimes a vector (as in the examples above), but sometimes also a data frame (e.g., when extracting two columns):

```{r}
airquality[, 1:2]
air[, 1:2]
```

If you want a vector for a single column, you can use `$` or `[[]]` for both data types (the latter is recommended).
