---
title: "3 – Solutions"
subtitle: "Statistical Data Analysis with R"
author: "Clemens Brunner"
date: 2024-10-17
format:
  html:
    page-layout: full
engine: knitr
knitr:
  opts_chunk: 
    R.options:
      width: 120
highlight-style: github
title-block-banner: true
theme:
  light: flatly
  dark: darkly
---

## Exercise 1

```{r}
r = 5  # radius
h = 9  # height
(A = 2 * r * pi * (r + h))  # area
(V = r**2 * pi * h)  # volume
```


## Exercise 2

```{r}
(x = c(4, 18, -7, 16, 4, -44))
(y = x**2)
(z = c(x, y))
length(z)
```


## Exercise 3

```{r}
x = c(44, 23, -56, 98, 99, 32, 45, 22)
x %% 2 == 0  # logical vector, TRUE for even elements
x[x %% 2 == 0]  # even elements of x
x %% 2 != 0  # logical vector, TRUE for odd elements
x[x %% 2 != 0]  # odd elements of x
```


## Exercise 4

```{r}
15:40
seq(75, 61, -3)
seq(14, 15, length.out=35)
```


## Exercise 5

```{r}
rep(c("Placebo", "Gruppe 1", "Gruppe 2"), each=10)
```


## Exercise 6

```{r}
k = seq(0, 20, 2)
k[-c(3, 7)]
k[1:5]
k[c(2, 5, 16)]  # element 16 does not exist -> NA
k[k > 11]
```


## Exercise 7

```{r}
t = c(10, 20, NA, 30, 40)
mean(t)  # result is NA
mean(t, na.rm=TRUE)  # ignores NA elements
```


## Exercise 8

```{r}
s = c(1, 11.3, 7.8, 3.4, 6)  # standard deviations
s**2  # variances
```


### Exercise 9

```{r}
x = c(2, 0, -5, 0, 1, -1, 0, 3, 0, 0, 7)
x == 0
sum(x == 0)
which(x == 0)
```


### Exercise 10

The indices inside the square brackets must always be contained in a single vector. If you want to index three positions, you need to create this vector with `c()`. Therefore, `x[2, 4, 6]` does not work because three vectors are specified (separated by commas) instead of a single vector.

The correct way to index the second, fourth, and sixth element of `x` is:

```{r}
x = 1:10
x[c(2, 4, 6)]
```
