---
title: "11 – Solutions"
subtitle: "Statistical Data Analysis with R"
author: "Clemens Brunner"
date: 2025-01-09
format:
  html:
    page-layout: full
engine: knitr
knitr:
  opts_chunk: 
    R.options:
      width: 120
highlight-style: github
title-block-banner: true
theme:
  light: flatly
  dark: darkly
---

## Exercise 1

### Height

```{r}
#| message: false
library(dplyr)

t.test(height ~ gender, data=starwars, alternative="less")
```

Female characters are not significantly shorter than male characters ($t=-1.25$, $p = .110$).


### Weight

```{r}
t.test(mass ~ gender, data=starwars, alternative="less")
```

Female characters are significantly lighter than male characters ($t=-1.87$, $p = .034$).


## Exercise 2

```{r}
library(palmerpenguins)
```


### Bill length

```{r}
with(
    subset(penguins, species == "Adelie" | species == "Chinstrap"),
    t.test(bill_length_mm ~ species)
)
with(
    subset(penguins, species == "Adelie" | species == "Gentoo"),
    t.test(bill_length_mm ~ species)
)
with(
    subset(penguins, species == "Gentoo" | species == "Chinstrap"),
    t.test(bill_length_mm ~ species)
)
```


### Bill depth

```{r}
with(
    subset(penguins, species == "Adelie" | species == "Chinstrap"),
    t.test(bill_depth_mm ~ species)
)
with(
    subset(penguins, species == "Adelie" | species == "Gentoo"),
    t.test(bill_depth_mm ~ species)
)
with(
    subset(penguins, species == "Gentoo" | species == "Chinstrap"),
    t.test(bill_depth_mm ~ species)
)
```

### Mit der Funktion `pairwise.t.test()`

```{r}
pairwise.t.test(penguins$bill_length_mm, penguins$species)
pairwise.t.test(penguins$bill_depth_mm, penguins$species)
```


## Exercise 3

For a paired *t*-test, we first need to convert the data to wide format:

```{r}
library(tidyr)

sleep_wide = pivot_wider(
    data=sleep,
    id_cols=ID,
    names_from=group,
    names_prefix="g",
    values_from=extra
)
```

Now we can perform the test:

```{r}
t.test(sleep_wide$g1, sleep_wide$g2, paired=TRUE)
```

The mean difference is −1.58, which is significantly different from zero ($p = .003$). The 95% confidence interval ranges from −2.46 to −0.7.
