---
title: "9 – Solutions"
subtitle: "Statistical Data Analysis with R"
author: "Clemens Brunner"
date: 2024-12-05
format:
  html:
    page-layout: full
engine: knitr
knitr:
  opts_chunk: 
    R.options:
      width: 120
highlight-style: github
title-block-banner: true
theme:
  light: flatly
  dark: darkly
---

## Exercise 1

By trying different values for $N$ we find $N = 62$ for $p < 0.05$. This means that a correlation of $r = 0.25$ is significant for $N = 62$ or larger.

```{r}
r = 0.25
N = 62
alpha = 0.05
z = atanh(r)
se_z = 1 / sqrt(N - 3)
(p = 2 * (1 - pnorm(z / se_z)))
```

Again, by trying different values for $N$ we find $N = 1538$ (at least) for $p < 0.05$. This shows that even a very small correlation becomes significant for a certain sample size.

```{r}
r = 0.05
N = 1538
alpha = 0.05
z = atanh(r)
se_z = 1 / sqrt(N - 3)
(p = 2 * (1 - pnorm(z / se_z)))
```


## Exercise 2

```{r}
x = c(8, 1, -4, 5, 6, 10, 9)
y = c(-2, -5, -6, 0, 3, 7, 10)
cor.test(x, y, conf.level=0.99)
```

The Pearson correlation is 0.8166, the corresponding $p$-value is 0.02501 (this is not significant for $\alpha=0.99$). The 99% confidence interval is $(-0.1405326, 0.9847508)$. It includes the value 0, which also indicates that the correlation is not significant.


## Exercise 3

```{r}
cor(mtcars[, c("mpg", "disp", "hp")], method="pearson")
cor(mtcars[, c("mpg", "disp", "hp")], method="spearman")
```

```{r}
pairs(subset(mtcars, select=c(mpg, disp, hp)))
```


## Exercise 4

```{r}
#| message: false
library(readr)
pm10 = read_csv("pm10.csv")
pm10$Datum = as.Date(pm10$Datum, format="%d.%m.%y")
```


### Time course of PM10 for both measurement locations

```{r}
plot(pm10$Datum, pm10$Petersgasse, type="l", col="blue", xlab="", ylab="PM10")
lines(pm10$Datum, pm10$DonBosco, col="red")
```


### Correlation between the two measurement locations

```{r}
with(pm10, plot(Petersgasse, DonBosco, pch=16, col=rgb(0, 0, 0, 0.5)))
abline(lm(pm10$DonBosco ~ pm10$Petersgasse), col="blue")

(r = cor.test(pm10$Petersgasse, pm10$DonBosco))
r$estimate**2  # explained variance

# we must exclude missing values explicitly
cor(pm10$Petersgasse, pm10$DonBosco, use="complete.obs")
```


## Exercise 5

```{r}
library(palmerpenguins)
cor(penguins$bill_length_mm, penguins$bill_depth_mm, use="complete.obs")
by(
    penguins[, c("bill_length_mm", "bill_depth_mm")],
    penguins$species,
    cor,
    use="complete.obs"
)
```
