---
title: "1 – Solutions"
subtitle: "Statistical Data Analysis with R"
author: "Clemens Brunner"
date: 2024-10-03
format:
  html:
    page-layout: full
engine: knitr
knitr:
  opts_chunk: 
    R.options:
      width: 120
highlight-style: github
title-block-banner: true
theme:
  light: flatly
  dark: darkly
---

## Exercise 1

The R version is displayed in the Console as an introductory message immediately after startup. The full version information is:

```
R version 4.4.1 (2024-06-14) -- "Race for Your Life"
```

Additionally, the R version is always visible in the title bar of the Console.

The RStudio version is 2024.04.2 Build 764, which can be found in the Help window, accessible through "Help" – "About RStudio".


## Exercise 2

```{r}
4 * pi * 6371^2
```


## Exercise 3

```{r}
(11 + 27 + 15 + 10 + 33 + 18 + 25 + 22 + 39 + 11) / 10
(11 * 27 * 15 * 10 * 33 * 18 * 25 * 22 * 39 * 11)^(1/10)
```


## Exercise 4

```{r}
sqrt(2) * (5^5 - pi) * 18 / ((7/5 + 13.2) * 7^(2/3))
```


## Exercise 5

The code doesn't work because the object `PI` does not exist (`Error: object 'PI' not found`). Case sensitivity matters, and the correct name for $\pi$ is `pi` (in lowercase):

```{r}
2 * 1 * pi
```
